The Evolution of Phenotypically Invariant Genetic Networks

### 2018-01-20 kalman_walk_paper.tex
edits

% wright1932roles mentions possibility of equivalent combinations
% wright1934evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model 
% fraisse2016genetics looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations
% chevin2010niche looks at incompats after adaptation in Fisher's geom model and how it depends on dimension
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
This has been analyzed for traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1934evolution,barton1986maintenance,barton2001role}
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
\citet{phillips1996maintenance} showed that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes.
Recent work has shown that Fisher's geometric model of stabilizing selection
predicts many general empirical patterns \citep{fraisse2016genetics,simon2017coadapted}.
We use similar tools in the second half of this paper,
differing in that we have a submanifold of optima rather than a single point.
Using simulations, \citet{tulchinsky2014hybrid}
%(and similarly \citet{khatri2015simple})
looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
if dosage compensation upregulates heterogametes by a factor of 2 relative to homogametes,
% as in \emph{Drosophila}
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
\plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

Above we synthesize concepts and tools from molecular and evolutionary biology with tools from control theory
to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection.
This model allows us to analytically describe the processes discussed verbally as
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical}
gives an analytical description of all phenotypically equivalent gene networks.
It also implies that nearly all systems are nonidentifiable,
and hence the general existence of axes of genetic variation that are not constrained by selection.
The independent movement of separated populations along these axes
can lead to reduction in hybrid viability, and hence speciation,
at a speed that depends on effective population size and amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and possibly rapid -- driver of speciation.
This may at first be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms.

Consistent with empirical observation of hybrid breakdown \citep[e.g.][]{plotner2017chlorosis}, 
we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
%\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids.
\plr{cite fisher's model stuff}

@@

\paragraph{On the origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probably important
% but this one may be strong since happens in all networks at once
Although, as classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
\plr{what}
and under stabilizing selection \citep{fierst},
previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
(Also note that the mechanism underlying hybrid inviability here 
resembles the Dobzhansky-Muller model,
but is more similar to the ``pathway model'' of \citet{lindtke2015genetic}.)
Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
In contrast, the model we develop here suggests that even under strictly neutral conditions, 
system drift can lead to speciation, at a rate fast enough 
to play a substantial role in species formation across the tree of life. 
Our results show that hybrids, under neutral processes, break down as a function of genetic distance,
and so may, in part, explain broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation patterns observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).

All of these forces -- adaptive shifts, conflict, and network drift --
are plausible drivers of speciation, and may even interact.
For instance, reinforcement due to local adaptation could in some situations provide a selective pressure
that speeds up system drift.
It remains to be seen how the relative strengths of these forces compare.
Furthermore,
while the fitness consequences of incompatibility in any one given network may be small, 
the cumulative impact of system drift across the many different networks an organism relies on may be substantial.

Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}). If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.

\paragraph{Population isolates and genetic load}
Above we have mostly discussed the case of two separated populations of equal size.
In contrast, consider a small subset of a large population that is suddenly isolated.
It will have a large amount of genetic variance, thanks to its history,
but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
By this point, it could have accumulated substantial incompatibility with the parental population.
Similarly, in a species with substantial isolation by distance, geographically local systems drift
might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.

@@

After selection, we have reproduction:
suppose this occurs as in the infinitesimal model \citep{barton2016infinitesimal},
so that each offspring of parents with traits $x$ and $y$
is drawn independently from a Gaussian distribution 
with mean $(x+y)/2$ and covariance matrix $R$.
Here, $R$ is the contribution of ``segregation variance'',
i.e., random choices of parental alleles.
If $\widetilde \Sigma = (\Sigma^{-1} + L^T L)^{-1}$ 
is the covariance matrix of the parents post-selection,
then the distribution of offspring will again be Gaussian,
with mean equal to that of the parents
and covariance matrix $\widetilde \Sigma/2 + R$.

@@

The description above does not describe how \emph{two} diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
To get an idea of how this might change,
suppose that a trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
so that being homozygous for the $+$ allele contributes $+2x_i$ to the trait.
For simplicity, we will neglect the effects of selection.
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\begin{align*}
    m &= 2 \sum_i (2 p_i - 1) x_i \\
    s^2 &= 4 \sum_i p_i (1-p_i) x_i^2 .
\end{align*}

Segregation variance between two parents depends on the loci at which either are heterozygous,
and each locus contributes independently
since alleles are additive.
If the alleles are at Hardy-Weinberg proportions,
then since segregation acts like a fair coin flip, a heterozygous locus contributes $x_i^2/4$ to the variance,
and so the \emph{mean} segregation variance, averaging across parents, is
% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)x^2.
\begin{align*}
    R_0(p) = \sum_i p_i (1-p_i) x_i^2 .
\end{align*}

On the other hand,
if the second parent came from a distinct population
with frequencies $q_i$
(an $F_1$ hybrid),
this would be
\begin{align*}
    R_1(p,q) &= \frac{1}{2} \sum_i p_i^2 (1-p_i)^2 x_i^2 
                + \frac{1}{2} \sum_i q_i^2 (1-q_i)^2 x_i^2 \\
                &= (R_0(p) + R_0(q))/2 .
\end{align*}
If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
and so $R_1(p,q) \approx R_0(p)$.

Now consider an $F_2$ hybrid, where both parents are $F_1$
and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
Then
\begin{align*}
    R_2(p,q) &= \frac{1}{2} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
\end{align*}
Suppose that the two populations are slightly drifted from each other,
with frequency difference $p_i-q_i = 2\epsilon_i$.
Then,
\begin{align*}
    p (1-q) + p (1-q)
    &=
    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
    &=
    2 u (1-u)
    % + \epsilon (u + (1-u) - u - (1-u))
    + 2 \epsilon^2 .
\end{align*}
If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
then $\epsilon$ has mean zero and variance roughly $2 u (1-u) t/N$.
Still assuming the populations are at stationarity,
so that $R_0$ is constant between the two,
and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
this implies that we expect
\begin{align*}
    R_2 
    &\approx 
    R_0 + 2 \sum_i p_i (1-p_i) x_i^2 t/N \\
    &=
    \left( 1 + \frac{2t}{N} \right) R_0 .
\end{align*}
On the other hand, the expected squared difference in trait \emph{means} between two such populations is
\begin{align} \label{eqn:var_dx}
    8 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
\end{align}
This implies that under this model,
segregation variance in $F_2$ hybrids between two populations
is roughly increased by a factor of $1/8$ of the difference between their means.

>>>>

% wright1932roles mentions possibility of equivalent combinations
% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model 
% fraisse2016genetics looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations
% chevin2010niche looks at incompats after adaptation in Fisher's geom model and how it depends on dimension
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
This has been analyzed for traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton2001role}
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
\citet{phillips1996maintenance} showed that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes.
Recent work has shown that Fisher's geometric model of stabilizing selection
predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
We use similar tools in the second half of this paper,
differing in that we have a submanifold of optima rather than a single point.
Using simulations, \citet{tulchinsky2014hybrid}
%(and similarly \citet{khatri2015simple})
looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
if dosage compensation upregulates heterogametes by a factor of 2 relative to homogametes,
% as in \emph{Drosophila}
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
Although many alleles will be dominant if the phenotype--fitness relationship is convex,
the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
instead, it derives from the nature of segregation load.
% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
% \plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

In this paper, we have used tools from quantitative genetics and control theory
to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection,
providing an explicit model of
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical}
gives an analytical description of all phenotypically equivalent gene networks.
It also implies that nearly all systems are nonidentifiable,
and hence the general existence of axes of genetic variation that are not constrained by selection.
The independent movement of separated populations along these axes by genetic drift
can lead to reduction in hybrid viability, and hence speciation,
at a speed that depends on effective population size and amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and possibly rapid -- driver of speciation.
This may at first be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms.

Consistent with empirical observation of hybrid breakdown \citep[e.g.][]{plotner2017chlorosis}, 
we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
%\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids,
as a result of segregation load \citep[similarly to][]{fraisse2016genetics}.

@@

\paragraph{The origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probably important
% but this one may be strong since happens in all networks at once
As classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms have been proposed
whereby Dobzhansky--Muller incompatibilities could originate
under neutral genetic drift \citep{lynch2000origin}
or stabilizing selection \citep{fierst}.
The same holds for the ``pathway model'' of \citet{lindtke2015genetic},
which is closer to the situation here.
However, previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
% Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
% In light of the few known incompatibilities, 
This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual},
two processes that typically act much more rapidly than genetic drift.
% In contrast, the model we develop here suggests that even under strictly neutral conditions, 
% system drift can lead to speciation, at a rate fast enough 
% to play a substantial role in species formation across the tree of life. 
However, our calculations suggest that even under strictly neutral processes,
hybrid fitness breaks down as a function of genetic distance
rapidly enough to play a substantial role in species formation across the tree of life. 
This is consistent with broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation rates observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).

All of these forces -- adaptive shifts, conflict, and network drift --
are plausible drivers of speciation, and may even interact.
Many of our observations carry over to models of directional selection --
for instance, rapid drift along the set of equivalent systems
could be driven by adaptation in a different, pleiotropically coupled system.
Or, reinforcement due to local adaptation might provide a selective pressure
that speeds up system drift.
Furthermore,
while the fitness consequences of incompatibility in any one given network may be small, 
the cumulative impact of system drift across the many different networks an organism relies on may be substantial. 
It remains to be seen how the relative strengths of these forces compare.

% Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}). If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.

% \paragraph{Population isolates and genetic load}
% Above we have mostly discussed the case of two separated populations of equal size.
% In contrast, consider a small subset of a large population that is suddenly isolated.
% It will have a large amount of genetic variance, thanks to its history,
% but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
% By this point, it could have accumulated substantial incompatibility with the parental population.
% Similarly, in a species with substantial isolation by distance, geographically local systems drift
% might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.

\paragraph{Fisher's geometric model}
There has been substantial work on models of quadratic stabilizing selection
around a single optimum -- i.e., Fisher's geometric model \citep[e.g.][]{simon2017coadapted}.
Our model would have this form if viewed only in phenotype space, 
where there is a single optimum --
but differs by explicitly considering evolutionary dynamics of underlying genotype-derived regulatory matrices.
These make the quantitative genetic model we are led to
a generalization of Fisher's geometric model,
with distinct behavior,
due to the existence of neutral directions in genotype space,
and curvature of the optimal set,
However, there are many shared aspects:
for instance, \citet{barton1989divergence} also found incompatibilities accumulating
with genetic drift.

@@

After selection, we have reproduction:
suppose this occurs as in the infinitesimal model \citep{barton2017infinitesimal},
so that each offspring of parents with traits $x$ and $y$
is drawn independently from a Gaussian distribution 
with mean $(x+y)/2$ and covariance matrix $R$.
Here, $R$ is the contribution of ``segregation variance'',
i.e., random choices of parental alleles.
If $\widetilde \Sigma = (\Sigma^{-1} + L^T L)^{-1}$ 
is the covariance matrix of the parents post-selection,
then the distribution of offspring will again be Gaussian,
with mean equal to that of the parents
and covariance matrix $\widetilde \Sigma/2 + R$.

@@

The description above does not describe how \emph{two} diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
To get an idea of how this might change,
suppose that a trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm \alpha_i$,
so that being homozygous for the $+$ allele contributes $+2 \alpha_i$ to the trait.
For simplicity, we will neglect the effects of selection.
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\begin{align*}
    m &= 2 \sum_i (2 p_i - 1) \alpha_i \\
    s^2 &= 4 \sum_i p_i (1-p_i) \alpha_i^2 .
\end{align*}

Segregation variance between two parents depends on the loci at which either are heterozygous,
and each locus contributes independently
since alleles are additive.
If the alleles are at Hardy-Weinberg proportions,
then since segregation acts like a fair coin flip, a heterozygous locus contributes $\alpha_i^2/4$ to the variance,
and so the \emph{mean} segregation variance, averaging across parents, is
% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)\alpha^2.
\begin{align*}
    R_0(p) = \sum_i p_i (1-p_i) \alpha_i^2 .
\end{align*}

On the other hand,
if the second parent came from a distinct population
with frequencies $q_i$
(an $F_1$ hybrid),
this would be
\begin{align*}
    R_1(p,q) &= \frac{1}{2} \sum_i p_i^2 (1-p_i)^2 \alpha_i^2 
                + \frac{1}{2} \sum_i q_i^2 (1-q_i)^2 \alpha_i^2 \\
                &= (R_0(p) + R_0(q))/2 .
\end{align*}
If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
and so $R_1(p,q) \approx R_0(p)$.

Now consider an $F_2$ hybrid, where both parents are $F_1$
and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
Then
\begin{align*}
    R_2(p,q) &= \frac{1}{2} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) \alpha_i^2  .
\end{align*}
Suppose that the two populations are slightly drifted from each other,
with frequency difference $p_i-q_i = 2\epsilon_i$.
Then,
\begin{align*}
    p (1-q) + p (1-q)
    &=
    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
    &=
    2 u (1-u)
    % + \epsilon (u + (1-u) - u - (1-u))
    + 2 \epsilon^2 .
\end{align*}
If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
then $\epsilon$ has mean zero and variance roughly $2 u (1-u) t/N$.
Still assuming the populations are at stationarity,
so that $R_0$ is constant between the two,
and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
this implies that we expect
\begin{align*}
    R_2 
    &\approx 
    R_0 + 2 \sum_i p_i (1-p_i) \alpha_i^2 t/N \\
    &=
    \left( 1 + \frac{2t}{N} \right) R_0 .
\end{align*}
On the other hand, the expected squared difference in trait \emph{means} between two such populations is
\begin{align} \label{eqn:var_dx}
    8 \sum_i p_i (1-p_i) \alpha_i^2 t / N = 8 R_0 t /N .
\end{align}
This implies that under this model,
segregation variance in $F_2$ hybrids between two populations
is roughly increased by a factor of $1/8$ of the difference between their means.
